<link href="http://kevinburke.bitbucket.org/markdowncss/markdown.css" rel="stylesheet"></link>

Day 3
====

Review
------
Assignment review: 
 * Relationship of students to teachers. Zonal differences in pupil to teacher ratio X 2. Just average number of students in primary schools.
 * Review "quiz": create a column that is 


Data cleaning:
----

We will be working mostly with our previous __sample_data__ set - so be sure that you have it loaded in your workspace. 

### Type conversion
Type conversion can be forced by `as.*` functions. Common __\*__ types you'd encounter are: 
  1. numeric
  2. integer
  3. character
  4. logical  
* Sometimes you'll encounter __factor__ variables, we recommend using __as.character()__ function to convert it into character type before proceeding 

```{r}
my_numbers <- c("1", "2", "3", "4", "TRUE")
my_numbers
as.numeric(my_numbers)
as.logical(my_numbers)
```


Data Cleaning
--------------
### string manipulations
* although regex suite comes with base R, string_r packages have similar implementations.
* we're not going to cover regex i.e. regular expression in this tuorial.
* getting position of matched pattern with grep()
```{r}
my_strings = c("Hello", "World", 'Foo')
grep(pattern="l", x=my_strings, ignore.case=FALSE)
#when value argument is set to true, grep() returns the actual strings matchs the patterns
grep(pattern="l", x=my_strings, ignore.case=FALSE, value=T)
```
* find pattern in strings and replace with sub()
* gsub() means global sub, which replace all the occurance of matching pattern, while sub() only works on the first appearance.
```{r}
my_strings
sub(pattern="o", replacement="X", x=my_strings)
gsub(pattern="o", replacement="X", x=my_strings)
```
* change to upper/lower case with toupper()/ tolower()
```{r}
my_strings
toupper(my_strings)
tolower(my_strings)
```
* concatenate strings with paste():
```{r}
paste("hello", "world", "foo", sep=",")
```
* if you're trying to concate every element in a vector/ list, use __collapse__ argument in paste()
```{r}
paste(my_strings, sep=',')
paste(my_strings, collapse=',')
```

### writing out data
* R can output multiple different output format, but we're going to cover csv and RDS format
* writing csv file with write.csv()
 * if you find output row.names everytime annoying, set row.names argument to FALSE in write.csv()
```{r}
write.csv(sample_data, "./my_output.csv", row.names=FALSE)
```
* we recommend output your data as RDS file, if you're only going to reuse it in R.
* use saveRDS() to save object in workspace to your harddrive
  * no row.names argument needs to be used, since we're saving the R object into a binary file
```{r}
saveRDS(sample_data, "./my_output.RDS")
```
* use readRDS() function to load saved RDS file
```{r echo=TRUE, eval=FALSE}
readRDS("./my_output.RDS")
```
```{r echo=FALSE, eval=TRUE}
readRDS("./my_output.RDS")[1:10, 1:10]
```


Advanced R
========================================================

creation of more complex columns(indicators) with __rowSums()__:
-------------------------------------
* column creation: sum of multiple numerical columns 
```{r}
sample_data$num_nurselabtechs_fulltime <- rowSums(cbind(sample_data$num_nurses_fulltime,
                                            sample_data$num_lab_techs_fulltime, na.rm = T))
```
  
```{r}
#now we can view all three variables: the new num_nurselabtechs_fulltime variable, and the two used to create it
head(subset(sample_data, select=c("num_nurses_fulltime", "num_lab_techs_fulltime", "num_nurselabtechs_fulltime")), 5)
```

creation of more complex columns(indicators) with __ifelse()__:
-------------------------------------


install packages from outside of cran
-------------------------------------
* in order to install packages on github we need some extra work
* this tutorial will use the example of formhub.R
* first step: install and load __devtools__ package from cran

```
install.packages('devtools') 
library(devtools)
```
* second step: use __install_github("repo_name", "user_name")__ function to install packages on github

```
install_github("formhub.R", username="SEL-Columbia")
library(formhub)
```

map functions: apply()
-----------------------
* instead of using for loops, we encourage using map. The apply/sapply is the implementation of map in R
* Here's the simply example for using apply to loop through every column and see the class of the column
  * note: __MARGIN = 2__ specifies doing column-wise process
```{r echo=TRUE, eval=FALSE}
apply(sample_data, MARGIN=2, FUN=class)
```
```{r echo=FALSE, eval=TRUE}
apply(sample_data, MARGIN=2, FUN=class)[1:10]
```

* you can also define your own function in apply()
* the following code returns sum of __NAs__ in each row
  * note: when you're define your own function in apply use semi-colon as the line marker.
  * note: __MARGIN = 1__ specifies doing row-wise process
```{r echo=TRUE, eval=FALSE}
apply(sample_data, MARGIN=1, function(x) {
  na_idx <- is.na(x);
  length(which(na_idx))
})
```
```{r echo=FALSE, eval=TRUE}
apply(sample_data, MARGIN=1, function(x) {
  na_idx <- is.na(x);
  length(which(na_idx))
  })[1:10]
```

improvements:
-----------------
* load your own functions into workspace with source()
```{r}
source("./my_source_functions.R")
my_sum(1,2)
```

* optimize ddply with idata.frame()
 * idata.frame optimizes the computation speed but at the cost of a slight more complicated code





