<link href="http://kevinburke.bitbucket.org/markdowncss/markdown.css" rel="stylesheet"></link>

Day 6
====


#### Creation of More Complex Columns(indicators) with __rowSums()__:
column creation: sum of multiple numerical columns 
```{r cache=T}
sample_data$num_nurselabtechs_fulltime <- rowSums(cbind(sample_data$num_nurses_fulltime,
                                            sample_data$num_lab_techs_fulltime, na.rm = T))
```
  
```{r cache=T}
#now we can view all three variables: the new num_nurselabtechs_fulltime variable, and the two used to create it
head(subset(sample_data, select=c("num_nurses_fulltime", "num_lab_techs_fulltime", "num_nurselabtechs_fulltime")), 5)
```

#### Creation of more complex columns(indicators) with __ifelse()__:




#### Data Pipeline
`outlier_normalized_health.R`, 
`outlier_normalized_education.R`
  - `outlier_functions.R`


`nmis_indicators_health_lga_level_normalized.R`, 
`nmis_indicators_education_lga_level_normalized.R`, 
`nmis_indicators_water_lga_level_normalized.R`

`nmis_indicators_health_facility_level_normalized.R`, 
`nmis_indicators_education_facility_level_normalized.R`,
`nmis_indicators_water_facility_level_normalized.R`

`nmis_indicators_COMBINING_normalized.R`

`nmis_post_processing`
  - `post_processing_functions.R`


