
Advanced R
========================================================

install packages from outside of cran
-------------------------------------
* in order to install packages on github we need some extra work
* this tutorial will use the example of formhub.R
* first step: install and load __devtools__ package from cran

```
install.packages('devtools') 
library(devtools)
```
* second step: use __install_github("repo_name", "user_name")__ function to install packages on github

```
install_github("formhub.R", username="SEL-Columbia")
library(formhub)
```

map functions: apply()
-----------------------
* instead of using for loops, we encourage using map. The apply/sapply is the implementation of map in R
* Here's the simply example for using apply to loop through every column and see the class of the column
  * note: __MARGIN = 2__ specifies doing column-wise process
```{r cache=T, echo=TRUE, eval=FALSE}
apply(sample_data, MARGIN=2, FUN=class)
```
```{r cache=T, echo=FALSE, eval=TRUE}
apply(sample_data, MARGIN=2, FUN=class)[1:10]
```

* you can also define your own function in apply()
* the following code returns sum of __NAs__ in each row
  * note: when you're define your own function in apply use semi-colon as the line marker.
  * note: __MARGIN = 1__ specifies doing row-wise process
```{r cache=T, echo=TRUE, eval=FALSE}
apply(sample_data, MARGIN=1, function(x) {
  na_idx <- is.na(x);
  length(which(na_idx))
})
```
```{r cache=T, echo=FALSE, eval=TRUE}
apply(sample_data, MARGIN=1, function(x) {
  na_idx <- is.na(x);
  length(which(na_idx))
  })[1:10]
```

improvements:
-----------------
* load your own functions into workspace with source()
```{r cache=T}
source("./my_source_functions.R")
my_sum(1,2)
```

* optimize ddply with idata.frame()
 * idata.frame optimizes the computation speed but at the cost of a slight more complicated code





